Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-09-21 bmc_template/manuscript.tex
tweak

%% fig 1 goes somewhere here
The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states (Figure~1). Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states) whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).

@@

Finally, if our goal is actually to estimate the number of discrete
behavioural modes underlying observed behaviour --- keeping in mind
that these discrete behavioural states are certainly an oversimplified
representation of animals' real internal states --- then, as we have
shown above, covariate-dependent models will generally be required to
avoid overestimating the number of states.  We note that researchers
in cluster analysis (of which HMMs are a special case) have shown that
the technical conditions required for BIC to apply may be violated
\cite{biernacki2000assessing}; however, BIC can be useful as an
approximate \emph{upper} limit on the number of states. Various
solutions to this problem have been proposed, including the
``integrated classification likelihood'' (ICL)
\cite{biernacki2000assessing}, as well as a simpler ``knee point''
method \cite{zhao_knee_2008} that looks for the cluster size that
corresponds to the largest change in BIC rather than to the smallest
overall BIC.  \cite{dean2012behavioural} took a similar approach, but
based on the log-likelihood curve rather than the BIC. However, HMMs are 
a special case even for the various solutions. ICL uses the likelihood of the 
most probable state sequence via the Viterbi algorithm to account for relative 
state occupancies between hidden states as a penalizing factor \cite{celeux2008selecting}, but it does
not have the same properties as the overall unconditional likelihood, such as, 
monotonic increase in likelihood as the number of hidden states increase. Furthermore, 
both log-likelihood and BIC based ``knee point'' methods look at the changes correponding to
a ``linear'' increase in number of states rather than the non-linear increase in parameters
corresponding to the linear increase in states. 
Nevertheless, at least in our simulations the BIC does correctly identify
the number of states when appropriate heterogeneity is included
in the model; the best practical criterion for identifying 
numbers of states underlying animal movement data is an open question.

@@

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/r_msdlist-1}
  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition). \bmb{do I need to explain transformation of LN SD ?}}
      \end{figure}

>>>>

%% fig 1 goes somewhere here
The simulation experiment supports our hypothesis that homogeneous transition HMMs overestimate the number of hidden states when the model is misspecified (Figure~1). Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).

@@

Finally, if our goal is actually to estimate the number of discrete
behavioural modes underlying observed behaviour --- keeping in mind
that these discrete behavioural states are certainly an oversimplified
representation of animals' real internal states --- then, as we have
shown above, covariate-dependent models will generally be required to
avoid overestimating the number of states.  We note that researchers
in cluster analysis (of which HMMs are a special case) have shown that
the technical conditions required for BIC to apply may be violated
\cite{biernacki2000assessing}; however, BIC can be useful as an
approximate \emph{upper} limit on the number of states. Various
solutions to this problem have been proposed, including the
``integrated classification likelihood'' (ICL)
\cite{biernacki2000assessing,celeux2008selecting}, as well as a simpler ``knee point''
method \cite{zhao_knee_2008} that looks for the cluster size that
corresponds to the largest change in BIC rather than to the smallest
overall BIC.  Dean \emph{et al.} \cite{dean2012behavioural} took a similar approach, but
based on the log-likelihood curve rather than the BIC. However, HMMs are 
a special case even for the various solutions. 
% ICL uses the likelihood of the  most probable state sequence via the Viterbi algorithm to account for relative  state occupancies between hidden states as a penalizing factor \cite{celeux2008selecting}, but it does not have the same properties as the overall unconditional likelihood, such as,  monotonic increase in likelihood as the number of hidden states increase. Furthermore,  both log-likelihood and BIC based ``knee point'' methods look at the changes corresponding to a ``linear'' increase in number of states rather than the non-linear increase in parameters corresponding to the linear increase in states.

Nevertheless, in our simulations the BIC does correctly identify
the number of states when appropriate heterogeneity is included
in the model; the best practical criterion for identifying 
numbers of states underlying animal movement data remains an open question.

@@

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/r_msdlist-1}
  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
      \end{figure}
